---
title: "4a_exercise"
output: html_document
date: "2024-09-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(tidyverse)
library(broom)

```


#QUESTION 1. Wrangle the data 

Step 1.1
You’ll start by exploring the data in aau1043_dnm.csv. First, load this data into a tibble.


```{r}

#Reading in the 'aau1043_dnm.csv' file into my environment as 'dnm'
dnm <- read_csv(file = "/Users/cmdb/qbb2024-answers/day4/4_afternoon/aau1043_dnm.csv")

#Printing the dnm dataframe to cheeck it was loaded in correctly 
head(dnm)

```


Step 1.2
Use group_by() and summarize() to tabulate the number of paternally and maternally inherited DNMs in each proband. Note that the maternal versus paternal origin of the mutations are recorded in the column titled Phased_combined.


```{r}

##Grouping the $Proband_id and summarizing it for paternal, maternal inherited DNMs

#Grouping and summarize to save result into 'dnm_summary'
dnm_summary <- dnm %>%
  group_by(Proband_id) %>%
  summarize(
    n_paternal_dnm = sum(Phase_combined == "father", na.rm = TRUE),
    n_maternal_dnm = sum(Phase_combined == "mother", na.rm = TRUE)
  )

#Printing the result to see 
head(dnm_summary)

```


Step 1.3
Now, load the data from aau1043_parental_age.csv into a new dataframe.

```{r}

##aau1043_parental_age.csv 

#Reading in the 'aau1043_parental_age.csv' file into my environment as 'ages'
ages <- read_csv(file = "/Users/cmdb/qbb2024-answers/day4/4_afternoon/aau1043_parental_age.csv")

#Printing the ages data-frame to check it was loaded in correctly 
head(ages)

```


Step 1.4
You now have two dataframes with complementary information. It would be nice to have all of this in one dataframe. Use the left_join() function to combine your dataframe from step 2 with the dataframe you just created in step 3 based on the shared column Proband_id.

```{r}

#Using 'left_join()' to combine the data-frames together 
dnm_combined <- left_join(dnm_summary, ages, by = "Proband_id")


#Printing 'dnm_combined' to check that the left_join properly worked 
head(dnm_combined)

```



#QUESTION 2. Fit and interpret linear regression models 

Using the merged dataframe from the previous section, you will be exploring the relationships between different features of the data.

Step 2.1
First, you’re interested in exploring if there’s a relationship between the number of DNMs and parental age. Use ggplot2 to plot the following. All plots should be clearly labelled and easily interpretable.

1. the count of maternal de novo mutations vs. maternal age
2. the count of paternal de novo mutations vs. paternal age

```{r}

#Plot 1. the count of maternal de novo mutations vs. maternal age

ggplot(data = dnm_combined, 
       mapping = aes(x = Mother_age, 
                     y = n_maternal_dnm)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  xlab("Mother age (years)") +
  ylab("Number of maternal de novo mutations") +
  ggtitle("The number of maternal de novo mutations vs. maternal age") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 15)) 

```


```{r}

#Plot 2. the count of maternal de novo mutations vs. maternal age

ggplot(data = dnm_combined, 
       mapping = aes(x = Father_age, 
                     y = n_paternal_dnm)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  xlab("Father age (years)") +
  ylab("Numner of paternal de novo mutations") +
  ggtitle("The number of paternal de novo mutations vs. paternal age") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 15)) 

```


Step 2.2
Now that you’ve visualized these relationships, you’re curious whether they’re statistically significant. 

2.2A: Fit a linear regression model to the data using the lm() function.

```{r}

#Performing lm based on the variables used in the plot above and saving the summarized results in 'mom_lm'
mom_lm <- lm(data = dnm_combined, 
             formula = n_maternal_dnm ~ 1 + Mother_age) %>% 
               summary()

#Print the linear regression result on 'dnm_combined' for the 2 variables saved in 'mom_lm' 
mom_lm

```

2.2B: What is the “size” of this relationship? In your own words, what does this mean? Does this match what you observed in your plots in step 2.1?

  ANSWER: 2.5 is the size of this relationship. This means that for each unit of the mother's age,
  the number of de novo mutations increases by 0.37757. This does match the scatterplot figure drawn
  in step 2.1 because the intercept and slope seem to relatively match up to the numbers written
  here. 


2.2C: Is this relationship significant? How do you know? In your own words, what does this mean?

  ANSWER: This relationship is significant based on the p-value. The p-value is less than 2.2e-16
  and commonly used significance p-value comparisons are either 0.05 or 0.01. This extremely small
  value means that the relationship between the age of the mother and the number of de novo
  mutations occurring is highly unlikely to have occurred by random chance, aka leading us to reject
  the null hypothesis and accept that these 2 variables have a statistically significant
  relationship to each other.  
  


Step 2.3

2.3A: As before, fit a linear regression model, but this time to test for an association between paternal age and paternally inherited de novo mutations.


```{r}

#Performing lm based on the variables used in the plot above and saving the summarized results in 'dad_lm'
dad_lm <- lm(data = dnm_combined, 
             formula = n_paternal_dnm ~ 1 + Father_age) %>% 
               summary()

#Print the linear regression result on 'dnm_combined' for the 2 variables saved in 'mom_lm' 
dad_lm

```

2.3B: What is the “size” of this relationship? In your own words, what does this mean? Does this match what you observed in your plots in step 2.1?

  ANSWER: 10.32632 is the size of this relationship. This means that for each unit of the father's
  age, the number of de novo mutations increases by 1.35384. This does match the scatterplot figure
  drawn in step 2.1 because the intercept and slope seem to relatively match up to the numbers
  written here.   


2.3C: Is this relationship significant? How do you know? In your own words, what does this mean?

  ANSWER: This relationship is significant based on the p-value. The p-value is less than 2.2e-16
  and commonly used significance p-value comparisons are either 0.05 or 0.01. This extremely small
  value means that the relationship between the age of the father and the number of de novo
  mutations occurring is highly unlikely to have occurred by random chance, aka leading us to reject
  the null hypothesis and accept that these 2 variables have a statistically significant
  relationship to each other. 
  



Step 2.4
Using your results from step 2.3, predict the number of paternal DNMs for a proband with a father who was 50.5 years old at the proband’s time of birth. Record your answer and your work (i.e. how you got to that answer).

```{r}

#Equation used to predict the number of paternal DNMs for a proband with a father who was 50.5 years of age 
y = 50.5*1.35384 + 10.32632 

#Based on the linear regression slope equation: y = m*x + b
  #10.32632 = intercept (b)
  #50.5 = the age of the father given (x)
  #1.35384 = the slope (m)
  #y = result we are looking for 

#Print 'y'
y

#ANSWER: I predict that there will be 78.69524 paternal DNMs for a proband with a father who was 505.5 years old. 

```




Step 2.5

Next, you’re curious whether the number of paternally inherited DNMs match the number of maternally inherited DNMs. Plot the distribution of maternal DNMs per proband (as a histogram). In the same panel (i.e. the same set of axes) plot the distribution of paternal DNMs per proband. Make sure to make the histograms semi-transparent so you can see both distributions.


```{r}

ggplot() +
  geom_histogram(data = dnm_combined, aes(x = n_maternal_dnm, fill = "Maternal"), 
                 alpha = 0.5, color = "white", binwidth = 1) +
  geom_histogram(data = dnm_combined, aes(x = n_paternal_dnm, fill = "Paternal"), 
                 alpha = 0.5, color = "white", binwidth = 1) +
  xlab("Number of inherited DNMs") +
  ylab("Count") +
  ggtitle("Count comparison of the number of paternally vs maternally inherited DNMs") +
  scale_fill_manual(name = "Inheritance Type", values = c("Maternal" = "orchid2", "Paternal" = "skyblue2")) +
  theme_classic() +
  theme(legend.key.size = unit(0.2, "cm"))

```


Step 2.6

2.6A: Now that you’ve visualized this relationship, you want to test whether there is a significant difference between the number of maternally vs. paternally inherited DNMs per proband. What would be an appropriate statistical model to test this relationship? Fit this model to the data.


```{r}

#Statistical test = paired t-test -> looks bi-modal but actually is unimodal 
res1 <- t.test(dnm_combined$n_maternal_dnm, dnm_combined$n_paternal_dnm, paired = TRUE)

#Print the results saved in 'res'
res1

```


After performing your test, answer the following questions:

2.6B: What statistical test did you choose? Why?

  ANSWER: I chose the paired t-test for my statistical model because we are looking at paired observations.
  Each proband has 2 measured values: the number of maternally inherited DNMs and the number of paternally
  inherited DNMs. Since these values come from the same individual, they are not independent of each other. To
  account for this type of data, a paired t-test is best because it tests for differences in means between 2
  related data measurements for a shared sample entry. 

2.6C: Was your test result statistically significant? Interpret your result as it relates to the number of paternally and maternally inherited DNMs.

  ANSWER: My result was statistically significant because the p-value is less than 2.2e-16, which is much
  smaller than the commonly used significance p-value comparisons 0.05 or 0.01. The highly significant p-value
  means that the observed difference between the number of maternally and paternally inherited DNMs per
  proband is very unlikely to have occurred by chance. Ultimately, we can state that the number of de novo
  mutations inherited from one parent (either maternally or paternally) differs significantly from the number
  inherited from the other parent (again, either maternally or paternally). 
