---
title: "4m_exercise_continued"
output: html_document
date: "2024-09-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

PYTHON DICTIONARIES 

#Question 7 

Finally, you can visualize how variable each gene’s expression is. Download the data, load it into R, and create a violin plot of expression levels broken down by gene (ggplot2’s geom_violin()).

For categories, create a combination of tissue names and gene IDs (dplyr::mutate(Tissue_Gene=paste0(Tissue, " ", GeneID)))



```{r}

#Load tidyverse library into the current R environment 
library(tidyverse)

#Load the tsv file into the current R environment and save as 'dicts_expression'
dicts_expression <- read_tsv("dicts_expr.tsv")

#Print the dictionary expression tsv file to see that it is properly loaded 
head(dicts_expression)

```



```{r}

#Using the 'mutate' command, create a new category called 'Tissue_Gene' to combine tissue names + gene IDs and save it back into 'dict_expression'
dicts_expression <- dicts_expression %>% 
  mutate(Tissue_Gene = paste0(Tissue, " ", GeneID))

#Print 'dicts_expression$Tissue_Gene' to see if it worked 
head(dicts_expression$Tissue_Gene)

```



```{r}

#Had trouble with the 'Expr' column being characters so I manually changed it to numerical values 
dicts_expression$Expr <- as.numeric(dicts_expression$Expr)

#Using the 'mutate' command, create a new column called 'log_exprs' by taking the log of the expression value and pseudocount of 1 
dicts_expression <- dicts_expression %>% 
  mutate(log_exprs = log(Expr + 1))

#Print 'dicts_expression$log_exprs' to see if it worked 
head(dicts_expression$log_exprs)

```




```{r}

#Create a ggplot violin plot and flip the x, y coordinates for the new Tissue, Gene category and its correspodning log pseudocount +1 expression value 
ggplot(data = dicts_expression, 
       mapping = aes(x = Tissue_Gene, 
                     y = log_exprs, 
                     fill = Tissue_Gene)) +
  geom_violin() +
  coord_flip() +
  xlab("Corresponding geneID to tissues") +
  ylab("Log expression values") + 
  ggtitle("Visualizing the log expression values for each gene ID and its corresponding tissue") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 7)) + 
  theme(legend.position = "none")

```


Questions on the results: 

A. Given the tissue specificity and high expression level of these genes, are you surprised by the results?

  Answer: 
  
  Given the tissue specificity and high expression levels of these genes, I was somewhat surprised by the         high variability in the data. I initially expected the expression levels of each gene to be more
  clustered within each tissue type. However, the data revealed was much more spread out than I imagined, with
  the expression levels differing quite significantly (for example: the 3rd stomach gene ranges from 0 to
  greater than 10). Despite this, the high log expression values were not unexpected. Given that the raw
  expression values were already quite large (in the thousands range), the high log expression values align with
  my expectations. The log values simply reflect the large magnitude of gene expression in specific tissues,
  reinforcing that these genes are indeed highly expressed where they are active.


B. What tissue-specific differences do you see in expression variability? Speculate on why certain tissues show low variability while others show much higher expression variability.

  Answer: 
  
  Except for pancreas, it seems that all the other tissues (liver, lung, minor salivary gland, thyroid, testis,
  stomach, small intestine, prostrate, and pituitary) have high variability. From those with high, stomach,       small intestine, prostrate, and pituitary are extremely spread out. I feel that specialized organs like the
  pancreas, usually have a limited number of roles and functions in the body. In comparison, the small intestine
  or pituitary gland has much more complex and diverse functions, which probably results in this higher
  variablity in expression. 
